SHARP - II: Revealing a bias in observational measurements of dark matter substructure with gravitational lens flux ratios

### 2015-11-17 papers/B1555_EdgeOn/Resubmission/referee_response.txt

RESPONSE: The referee is correct to point out that we did not quantify
the bulge/disk ratio in the lensing galaxy.  We have performed a
bulge+disk fit to the lens-galaxy surface brightness distribution.
Our fit returns a bulge component that is consistent with zero
emission.  We have added text to the paper describing this result.  We
note that the two components in our lens model can be thought of as
the edge-on disk and the dark-matter halo of the lensing galaxy.

> More importantly, the large disk component in the lens model produces
> a significant distortion to the usually ellipsoidal critical curves
> found in strong lenses, as shown in Fig 3.  It has long been known
> that such departures from ellipsoidal density distributions are
> tightly constrained by extended emission from the source galaxies,
> e.g. arcs and Einstein rings that span a large range of azimuthal
> angle.  The imaging shown in Fig 2 exhibits faint extended arclets
> from the source galaxy that could constrain disky components in the
> lens mass distribution, but the lens modeling in the paper neglects to
> fit the extended emission, and instead only models the positions and
> fluxes of the pointlike radio quasar.  The signal-to-noise of the
> extended emission may be too low to provide meaningful constraints,
> but given that the authors have this data in hand, and given that the
> authors have considerable experience modeling this type of extended
> emission in lens systems, they should attempt to fit the observed
> arclets to see if a 10% mass fraction in an off-center thin disk is
> consistent with the observed arclets.

RESPONSE: The referee is correct in pointing out that the modelling of
the extended images would provide tighter constraints on the proposed
model. As a matter of fact, we had hoped to do exactly the analysis
that the referee has suggested.  Unfortunately, in the case of B1555
IR imaging the SNR of the lensed extended emission is low.  More
importantly, the disk is responsible for a significant level of dust
extinction that leave images D and B unobservable, and even the arc
near image A strongly affected. We have, however, tried modelling this
dataset by only focusing on image C which is not affected by dust, and
the results are consistent with those found by modelling the point
images positions and fluxes.  Similarly we hoped to use the VLBI
images for the extended source modeling, but as mentioned in the text
the poor quality of the data prevented us from modelling this dataset
in a reliable manner.  We have added text to the manuscript explaining
this.

@@

> 
> The final section of the paper speculates that massive disc components
> may be prevalent in strong lenses, and that such discs may bias
> inferences on the abundance of dark matter substructure using strong
> lensing.  The authors fail to provide evidence supporting these
> speculations, even though they have the data to do so.  As noted
> above, strong lenses with extended source emission (arcs and rings)
> tightly constrain the presence of thin components in lens galaxies.
> In some cases with Einstein rings, disc components are excluded from
> contributing more than percent-level fractions of the mass density.
> It would therefore be highly surprising for a substantial fraction of
> strong lenses to contain hidden, massive edge-on discs that could
> distort the lensing so significantly, as the paper intimates.
> 
> Rather than wildly speculating, the authors could easily check the
> plausibility of this claim.  A number of the authors were members of
> the SLACS collaboration, whose dataset contains many lenses with
> extended emission from the source galaxies.  These systems are the
> obvious way to quantify the fraction of lens systems containing
> massive edge-on discs, since the arcs in many SLACS lenses span a wide
> range of azimuthal angle.  Once they have quantified the abundance of
> massive edge-on discs, the authors should then perform Monte Carlo
> calculations of quasar lenses to calculate the statistics of flux
> ratio anomalies produced by these discs.  Only then can they state
> whether hidden edge-on massive discs can bias observational
> measurements of dark matter substructure with gravitational lens flux
> ratios.

RESPONSE: In the SHARP sample, we find that 5 out of the 25 systems
that we imaged have clear disk components, indicating that this is not
a completely unexpected phenomenon.  While we agree with the referee
that it is important to quantify how common disks are in lensing
galaxies, we do not feel that the SLACS sample provides the
appropriate comparison.  This is because a significant portion of the
SHARP systems come from source-selected samples, such as CLASS, that
are agnostic about the nature of the lensing galaxy.  In fact, CLASS
contains at least one other system, B1600+434 (albeit a double), where
the lensing galaxy appears to be an edge-on disk with a very low
bulge/disk ratio.  In contrast, SLACS lenses were selected to be
systems where the lensing galaxy was known a priori to be a massive
elliptical galaxy.  We would therefore expect a lower incidence of
disky systems in SLACS than in the SHARP sample.  We agree completely
with the referee that doing a quantification of the statistics of
massive disks and their effects on flux-ratio anomalies is an
extremely important task and, in fact, this is one of the future
projects that we are planning.  However, we feel that such an
extensive analysis is beyond the scope of this letter.  As such, we
have modified the text to focus more on this individual lens system
and put more caveats on the possible generalization of the results.

### 2015-11-17 papers/B1555_EdgeOn/B1555_letter.tex
More small changes.

{\bf Here, we find evidence for an edge-on disc in the flux-ratio
anomaly gravitational lens B1555+375 that can also account for
the required perturbation in the mass model. Such high-order deviations from a smooth mass distribution have already been suggested previously \citep{evans03, congdon05}, but this is the first time that an observed disc has been explicitly modelled to account for a flux-ratio anomaly. However, it is important to realise that by including observationally-motivated perturbations to a `smooth' mass distribution, like luminous dwarf companions or edge-on discs, in the mass models,
we are still limited in our interpretation since we must assume that
the anomalies are due exclusively to these additional components.
For B1555+375, it may be possible to independently constrain the
mass of the disc through kinematic measurements (most likely from
mm-wavelength resolved emission line observations), which would
test whether the disc makes a sufficient contribution to the mass
model (e.g. as in the case of B1933+503; \citet{suyu12}). Also, as
we have shown here, B1555+375 shows evidence for an extended $\sim$ 100 mas gravitational arc at radio wavelengths that crosses the
edge-on disc. Further high resolution imaging with very long baseline interferometry will provide the observational constraints required to fully test whether the disc is sufficient to cause the anomaly.}

\textbf{ Our analysis of B1555+375 demonstrates that the choice of the macromodel used to describe the `smooth' lens galaxy mass distribution, combined with the limited information provided by the position of compact lensed images, can lead to a misinterpretation of the origin of flux-ratio anomalies. Before concluding that the flux-ratio anomalies in any given lens system are the result of dark matter substructure, then, it is critical that other possible sources of significant curvature in the lensing potential (i.e., `sharp' mass features such as edge-on disks or even massive luminous satellites) are taken into account 
 (e.g., \citet{Xu15}). This result is timely, as upcoming large-scale surveys (e.g., the Dark Energy Survey, Kilo-Degree Survey, Euclid, and the Large Synoptic Survey Telescope) will deliver orders of magnitude more gravitationally lensed quasars for which flux ratios can
  be measured using narrow lines (e.g., \citet{N14}). Near-infrared AO imaging of other lenses in the SHARP sample shows that B1555+375 is not the only system having a significant disc component detection in the lensing galaxy; these gravitational lenses will be investigated in future work, but the images alone suggest that other systems may have similar links between flux-ratio anomalies and disc components. Finally, we note that massive edge-on discs are \textit{not} observed in some flux-ratio anomaly systems (and the presence of a disc-like component may not resolve the anomaly anyways; \citet{evans03}), indicating that these systems are still useful for probing the dark matter substructure mass function.}

@@

\section*{Acknowledgments}
We are grateful to Dandan Xu for useful discussions. CDF and DJL acknowledge support from NSF-AST-0909119.  LVEK is supported in part through an NWO-VICI career grant (project number 639.043.308). The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc. Based on observations made with the NASA/ESA Hubble Space Telescope, obtained from the data archive at the Space Telescope Science Institute. STScI is operated by the Association of Universities for Research in Astronomy, Inc. under NASA contract NAS 5-26555. Some of the data presented herein were obtained at the W. M. Keck Observatory, which is operated as a scientific partnership among the California Institute of Technology, the University of California and the National Aeronautics and Space Administration. The Observatory was made possible by the generous financial support of the W. M. Keck Foundation. The authors wish to recognize and acknowledge the very significant cultural role and reverence that the summit of Mauna Kea has always had within the indigenous Hawaiian community.  We are most fortunate to have the opportunity to conduct observations from this mountain.

>>>>

{\bf Here, we find evidence for an edge-on disc in the flux-ratio
  anomaly gravitational lens B1555+375 that can also account for the
  required perturbation in the mass model. Such high-order deviations
  from a smooth mass distribution have already been suggested
  previously \citep{evans03, congdon05}, but this is the first time
  that an observed disc has been explicitly modelled to account for a
  flux-ratio anomaly. However, it is important to realise that by
  including observationally-motivated perturbations to a `smooth' mass
  distribution, such as luminous dwarf companions or edge-on discs, in
  the mass models, we are still limited in our interpretation since we
  must assume that the anomalies are due exclusively to these
  additional components.  For B1555+375, it may be possible to
  independently constrain the mass of the disc through kinematic
  measurements (most likely from mm-wavelength resolved emission line
  observations), which would test whether the disc makes a sufficient
  contribution to the mass model (e.g. as in the case of B1933+503;
  \citet{suyu12}). Also, as we have shown here, B1555+375 shows
  evidence for an extended $\sim$ 100 mas gravitational arc at radio
  wavelengths that crosses the edge-on disc. Further high resolution
  imaging with very long baseline interferometry will provide the
  observational constraints required to fully test whether the disc is
  sufficient to cause the anomaly.}

\textbf{ Our analysis of B1555+375 demonstrates that the choice of the macromodel used to describe the `smooth' lens galaxy mass distribution, combined with the limited information provided by the position of compact lensed images, can lead to a misinterpretation of the origin of flux-ratio anomalies. Before concluding that the flux-ratio anomalies in any given lens system are the result of dark matter substructure, therefore, it is critical that other possible sources of significant curvature in the lensing potential (i.e., `sharp' mass features such as edge-on disks or even massive luminous satellites) are taken into account 
 (e.g., \citet{Xu15}). This result is timely, as upcoming large-scale surveys (e.g., the Dark Energy Survey, Kilo-Degree Survey, Euclid, and the Large Synoptic Survey Telescope) will deliver orders of magnitude more gravitationally lensed quasars for which flux ratios can
  be measured using narrow lines (e.g., \citet{N14}). Near-infrared AO imaging of other lenses in the SHARP sample shows that B1555+375 is not the only system having a significant disc component detection in the lensing galaxy; these gravitational lenses will be investigated in future work, but the images alone suggest that other systems may have similar links between flux-ratio anomalies and disc components. Finally, we note that massive edge-on discs are \textit{not} observed in some [most??] flux-ratio anomaly systems (and the presence of a disc-like component may not resolve the anomaly anyways; \citet{evans03}), indicating that these systems are still useful for probing the dark matter substructure mass function.}

@@

\section*{Acknowledgments}
We thank the referee for comments that improved the paper.  We are grateful to Dandan Xu for useful discussions. CDF and DJL acknowledge support from NSF-AST-0909119.  LVEK is supported in part through an NWO-VICI career grant (project number 639.043.308). The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc. Based on observations made with the NASA/ESA Hubble Space Telescope, obtained from the data archive at the Space Telescope Science Institute. STScI is operated by the Association of Universities for Research in Astronomy, Inc. under NASA contract NAS 5-26555. Some of the data presented herein were obtained at the W. M. Keck Observatory, which is operated as a scientific partnership among the California Institute of Technology, the University of California and the National Aeronautics and Space Administration. The Observatory was made possible by the generous financial support of the W. M. Keck Foundation. The authors wish to recognize and acknowledge the very significant cultural role and reverence that the summit of Mauna Kea has always had within the indigenous Hawaiian community.  We are most fortunate to have the opportunity to conduct observations from this mountain.
